---
title: "Classifying the Onset of Diabetes Using Medical and Demographic Patient Data"
output: 
  html_document:
    theme: journal
    toc: True
    toc_float: True
---

# Overview

## Setup Environment

<!--
Load necessary libraries - (run it on the console)
# Install packages
install.packages("caret")
install.packages("correlationfunnel")
install.packages("corrplot")
install.packages("DataExplorer")
install.packages("dplyr")
install.packages("forcats")
install.packages("GGally")
install.packages("ggthemes")
install.packages("ggplot2")
install.packages("naniar")
install.packages("summarytools")
install.packages("tidyverse")
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load packages
library(caret)
library(correlationfunnel)
library(corrplot)
library(DataExplorer)
library(dplyr)
library(forcats)
library(GGally)
library(ggthemes)
library(ggplot2)
library(naniar)
library(summarytools)
library(tidyverse)
```

## Dataset Overview and Objective

The **Diabetes Prediction** Dataset is a comprehensive collection of medical and demographic records aimed at identifying the likelihood of diabetes in patients. Each observation in the dataset includes both physiological and lifestyle-related features, along with a binary outcome indicating whether the individual is diabetic (positive) or not (negative).

Key variables in the dataset include:

Age: The age of the patient
Gender: Male or Female
Body Mass Index (BMI): A measure of body fat based on height and weight
Hypertension: Indicates whether the patient has high blood pressure (1 = Yes, 0 = No)
Heart Disease: Indicates the presence of cardiovascular disease (1 = Yes, 0 = No)
Smoking History: Categorical variable showing smoking habits (e.g., never, current, former)
HbA1c Level: Average blood sugar level over the past 2–3 months, measured in percentage
Blood Glucose Level: Instantaneous measure of blood sugar at the time of recording
Diabetes Status: Target variable indicating whether the patient is diabetic (positive) or not (negative)
The primary objective of this analysis is to develop a predictive classification model that determines whether an individual is likely to have diabetes based on their medical and lifestyle attributes. 

# Data Exploration

## Load Data

```{r Load Data}
# Clear everything
rm(list = ls())
gc()

# Load only essential packages
library(ggplot2)
library(dplyr)

# Load data
df <- read.csv("diabetes_prediction_dataset.csv", stringsAsFactors = FALSE)

# View first few rows
head(df)

# View structure and summary
summary(df)
```

```{r Data Glimpse, include=FALSE}
dplyr::glimpse(df)
```

## Data Preprocessing

### Missing Values

```{r Missing Values}
# Check for missing values
sum(is.na(df))
vis_miss(df)  # Visualize missing data
```

### Encode Categorical Variables

Categorical variables were converted to factors for better readability for EDA purpose. Binary values (0 and 1) for hypertension, heart disease, and diabetes were relabeled as "No"/"Yes" or "Negative"/"Positive" to improve interpretability.

```{r Data Encoding}
# Convert to proper types -> 0 and 1 to "yes" and "no"
df <- df %>%
  mutate(
    gender = as.factor(gender),
    smoking_history = as.factor(smoking_history),
    hypertension = factor(hypertension, levels = c(0,1), labels = c("No", "Yes")),
    heart_disease = factor(heart_disease, levels = c(0,1), labels = c("No", "Yes")),
    diabetes = factor(diabetes, levels = c(0,1), labels = c("Negative", "Positive"))
  )

# Check "No Info" in smoking history
df %>% 
  count(smoking_history) %>% 
  mutate(prop = n/sum(n))
```

### Feature Engineering

Smoking history was cleaned by renaming "No Info" to "Unknown" and merging similar categories. "Ever," "former," and "not current" became "Former" (19.8%), while "Never" (35.1%) and "Current" (9.3%) were kept unchanged.

```{r Feature Engineering}
# Consolidate smoking history categories
df <- df %>%
  mutate(smoking_history = factor(smoking_history)) %>%
  mutate(
    smoking_history = fct_collapse(smoking_history,
      "Never" = "never",
      "Former" = c("ever", "former", "not current"),
      "Current" = "current",
      "Unknown" = "No Info"
    )
  )

# Check "No Info" in smoking history
df%>% 
  count(smoking_history) %>% 
  mutate(prop = n/sum(n))
```

# Exploratory Data Analysis

## Diabetes (Response) Class Distribution

```{r Diabetes Class Distribution}
# Target distribution
ggplot(df, aes(diabetes, fill = diabetes)) +
  geom_bar() +
  geom_text(
    stat = "count", 
    aes(label = ..count..), 
    vjust = -0.5
  ) +
  labs(title = "Diabetes Class Distribution", x = "Diabetes Status")
```

## Diabetes Versus Age Distribution

```{r Diabetes vs Age}
# Age analysis
ggplot(df, aes(x = age, fill = diabetes, group = diabetes)) +
  geom_density(alpha = 0.7) +
  facet_wrap(~diabetes, ncol = 1) +
  labs(title = "Age Distribution by Diabetes Status", x = "Age", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Distribution of BMI

```{r BMI Distribution}
# BMI analysis with clinical thresholds
ggplot(df, aes(bmi)) +
  geom_histogram(fill = "steelblue", bins = 30) +
  geom_vline(xintercept = c(18.5, 25, 30), linetype = "dashed", color = "red") +
  annotate("text", 
           x = c(17, 21.5, 26.5, 32.5),  
           y = c(3000, 8000, 12000, 8000), 
           label = c("Underweight", "Normal", "Overweight", "Obese")) +
  labs(title = "BMI Distribution with Clinical Thresholds")
```

## Critical Biomarkers Versus Diabetes Class

```{r Biomarker Versus Diabetes Class}
df %>% 
  select(HbA1c_level, blood_glucose_level, diabetes) %>% 
  pivot_longer(-diabetes) %>% 
  ggplot(aes(x = value, fill = diabetes)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~name, scales = "free", ncol = 1) +
  geom_vline(data = data.frame(name = c("HbA1c_level", "blood_glucose_level"),
                               threshold = c(6.5, 200)),
             aes(xintercept = threshold), color = "red", linetype = "dashed") +
  labs(title = "Critical Biomarkers Distribution",
       subtitle = "Red lines indicate diagnostic thresholds")
```

## Diabetes Class Distribution by Gender

```{r Diabetes Class by Gender}
ggplot(df, aes(x = gender, fill = gender)) +
  geom_bar() +
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.5,
    size = 3
  ) +
  facet_wrap(~diabetes) +  # Add faceting by diabetes status
  labs(
    title = "Gender Distribution by Diabetes Status",
    x = "Gender",
    y = "Count"
  ) +
  theme_minimal()
```

We are dropping "others" because they are only  18 observations (0.018% of 100k dataset) won't impact analysis

## Distribution of Gender

```{r Gender Distribution}
table(df$gender)

df <- df %>%
  filter(gender %in% c("Male", "Female"))

#Drop unused factor levels
df$gender <- droplevels(df$gender)

# Create the plot
ggplot(df, aes(x = gender, fill = gender)) +
  geom_bar() +
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.5,
    size = 4,
    color = "black"
  ) +
  labs(
    title = "Gender Distribution (Male/Female only)",
    subtitle = "Excluded 18 observations of 'Other' gender",
    x = "Gender",
    y = "Count"
  ) +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
  theme_minimal()
```

## Correlation Funnel for Target-Focused Relationships

```{r Correlation Funnel}
# Correlation funnel for target-focused relationships

# Prepare data
df <- df%>%
  mutate(
    diabetes = factor(diabetes, levels = c("Negative", "Positive")),
    across(where(is.character), as.factor)
  )

# Create correlation funnel
df %>%
  binarize() %>%
  correlate(target = diabetes__Positive) %>% 
  plot_correlation_funnel(interactive = FALSE) +
  labs(title = "Diabetes Risk Correlation Funnel") +
  theme_minimal()
```

## Key Findings from Correlation Analysis

High-Risk Indicators
• HbA1c > 6.2% & glucose > 140 mg/dL
• Age > 60, hypertension, heart disease

Protective Factors
• HbA1c < 4.8% & glucose < 100 mg/dL
• BMI < 23.63, age < 24

Notable Patterns

Clear biomarker thresholds
Progressive age-risk relationship
Gender shows minor association

## Diabetes Rate by Comorbidities

```{r Diabetes Rate by Comorbidities}
# Comorbidity analysis
comorbidity_analysis <- df %>%
  group_by(hypertension, heart_disease) %>%
  summarise(
    diabetes_rate = mean(diabetes == "Positive"),
    count = n(),
    .groups = "drop"
  )

ggplot(comorbidity_analysis, aes(hypertension, heart_disease, fill = diabetes_rate)) +
  geom_tile() +
  geom_text(aes(label = scales::percent(diabetes_rate, accuracy = 0.1))) +
  scale_fill_gradient(low = "white", high = "firebrick") +
  labs(title = "Diabetes Rate by Comorbidities",
       x = "Hypertension", y = "Heart Disease")
```

## Data Preprocessing 


```{r}
# Handle BMI outliers using quantile capping
summary(df$bmi)

#calculate the 0.5% and 99.5% percentiles
bmi_caps <- quantile(df$bmi, probs = c(0.005, 0.995), na.rm = TRUE)
print(bmi_caps)
```
The observed BMI values, ranging from 13.89 to 52.50, fall within the clinically plausible range of 12 to 60. This suggests that all recorded measurements are physiologically valid. As a result, we chose to retain all BMI values without applying any capping.



```{r}
# Convert necessary variables to numeric for correlation
corr_data <- df %>%
  mutate(
    hypertension_num = as.numeric(hypertension) - 1,  
    heart_disease_num = as.numeric(heart_disease) - 1,
    diabetes_num = as.numeric(diabetes) - 1
  ) %>%
  select(age, bmi, HbA1c_level, blood_glucose_level, 
         hypertension_num, heart_disease_num, diabetes_num)

# Compute correlation matrix
cor_matrix <- cor(corr_data)
cor_matrix


```

```{r}
#correlation
corrplot(cor_matrix, 
         method = "color",        
         type = "upper",         
         tl.col = "black",       
         tl.srt = 45,           
         addCoef.col = "black",  
         number.cex = 0.7,       
         diag = FALSE)          

# Print notable correlations (> 0.3)
notable <- which(abs(cor_matrix) > 0.3 & upper.tri(cor_matrix), arr.ind = TRUE)

if(nrow(notable) > 0) {
  cat("\nNotable Correlations:\n")
  for(i in 1:nrow(notable)) {
    row <- notable[i,1]
    col <- notable[i,2]
    cat(rownames(cor_matrix)[row], "&", 
        colnames(cor_matrix)[col], ":", 
        round(cor_matrix[row,col], 2), "\n")
  }
} else {
  cat("\nNo notable correlations found (all r < 0.3)\n")
}
```
The correlation matrix reveals moderate relationships between key clinical variables and diabestes, so we won't change anything 

```{r}
# Create clinically relevant features
df <- df %>%
  mutate(
    high_hba1c = factor(ifelse(HbA1c_level >= 6.5, "Yes", "No")),
    high_glucose = factor(ifelse(blood_glucose_level >= 200, "Yes", "No")),
    age_group = cut(age, breaks = c(0, 30, 45, 60, 100),
                    labels = c("<30", "30-45", "46-60", "60+")),
    bmi_category = cut(bmi, breaks = c(0, 18.5, 25, 30, 100),
                      labels = c("Underweight", "Normal", "Overweight", "Obese"))
  )

head(df)

```



```{r}
# prepare modeling data
model_data <- df %>%
  select(-age_group, -bmi_category) %>%
  mutate(diabetes = as.factor(diabetes))

# create encoding model (EXCLUDING diabetes)
dummy_model <- dummyVars(~ . , 
                         data = model_data %>% select(-diabetes),
                         fullRank = TRUE)

# apply encoding to predictors only
encoded_predictors <- predict(dummy_model, newdata = model_data) %>% 
  as.data.frame()

# add target variable back
final_data <- encoded_predictors %>% 
  mutate(diabetes = model_data$diabetes)

```


```{r, include=FALSE}
glimpse(final_data)
```
```{r Save Final Data to CSV, include=FALSE}
# Save processed data
# write_csv(final_data, "diabetes_processed.csv")
```

Test/Train/Validation Splits

```{r}
set.seed(35)
split_df <- final_data

train_index <- createDataPartition(final_data$diabetes, p = 0.8, list = FALSE)
train_data <- final_data[train_index, ]
test_val_data <- final_data[-train_index, ]

test_val_index <- createDataPartition(test_val_data$diabetes, p = 0.5, list = FALSE)
validation_data <- test_val_data[test_val_index, ]
test_data <- test_val_data[-test_val_index, ]

dim(final_data)
dim(train_data)
dim(test_data)
dim(validation_data)
```

```{r}
ggplot(train_data, aes(diabetes, fill = diabetes)) +
  geom_bar() +
  geom_text(
    stat = "count", 
    aes(label = ..count..), 
    vjust = -0.5
  ) +
  labs(title = "Diabetes Class Distribution", x = "Diabetes Status")
```

```{r}
ggplot(test_data, aes(diabetes, fill = diabetes)) +
  geom_bar() +
  geom_text(
    stat = "count", 
    aes(label = ..count..), 
    vjust = -0.5
  ) +
  labs(title = "Diabetes Class Distribution", x = "Diabetes Status")
```

```{r}
ggplot(validation_data, aes(diabetes, fill = diabetes)) +
  geom_bar() +
  geom_text(
    stat = "count", 
    aes(label = ..count..), 
    vjust = -0.5
  ) +
  labs(title = "Diabetes Class Distribution", x = "Diabetes Status")
```

```{r, include=FALSE}
# write_csv(train_data, "diabetes_train.csv")
# write_csv(test_data, "diabetes_test.csv")
# write_csv(validation_data, "diabetes_validation.csv")
```

# Model Development

## Niat's Model

### Model Building

```{r}

```

### Model Testing and Evaluation

```{r}

```

## Rebecca's Model

### Model Building

```{r}

```

### Model Testing and Evaluation

```{r}

```

## Tommy's Model

### Model Building

```{r}

```

### Model Testing and Evaluation

```{r}

```